We Are All Trying to Describe Something And We Don't Have the Words

### 2023-04-19 outputs/paper/paper.qmd
fixed the bib

---
title: "We Are All Trying to Describe Something And We Don't Have the Words"
subtitle: "The Jack Antonoffification of Pop Music?"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/jack."
date: today
abstract: "Using data from the Spotify API, I explored Jack Antonoff's production style and assessed whether Spotify's audio features can capture it. Using a logistic regression model, I found that danceability was the most important feature in predicting Antonoff's involvement. In an interview with Glen McDonald, an engineer at Spotify, I learned that danceability and valence were created by tagging songs as positive or danceable. In the discussion, I do a close-listen to Antonoff's most danceable songs, and argue that what danceability is capturing the visceral quality of Antonoff's music -- which checks out with his critics' visceral reaction to his music."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Midnights was produced by Swift and her longtime collaborator Jack Antonoff, who is the subject of this paper, and, I suspect, the reason behind both my blacking out, and the stranger's "Exactly".

Antonoff is prolific, to the extent that it is starting to bother people. He has produced and co-produced music for Lorde, Taylor Swift, Florence and the Machine, The Chicks, Clairo, The 1975, Grimes, Zayn, Pink, Lana Del Rey, Olivia Rodrigo, the Minions: The Rise of Gru Soundtrack, and more. All of this and his own bands too (Bleachers, and fun.). When I was getting brunch with a friend one Sunday and a Swift song came on, and she said "Jack Antonoff's gonna kill us all", we both laughed. What my friend was referring to, of course, is the fact that Jack Antonoff and his sound seems to be everywhere these days. As Andrew Marantz writes in the New Yorker: "When his band releases an album, the world responds politely. When he produces one by Lorde or Lana Del Rey or Taylor Swift, the world wobbles on its axis" [@citeMarantz].

@@

I spoke to Caleb Gamman, a Youtuber and Antonoff-production-style hater who had a tweet go viral around the time of Taylor Swift's *Midnights* release, in which he predicts whether or not a song was produced by Antonoff within seconds. A real-life logistic regression, if you will. Using an actual logistic regression model, I was able to determine that Spotify's `danceability` feature was the most important in predicting whether or not a song had any Antonoff involvement. During an interview with Glen McDonald, an engineer at Spotify, I learned that the `valence` and `danceability` variables were created by hiring college interns to tag whether or not songs were positive or danceable. In the discussion section, I listen to the songs with the highest danceability scores, and make an argument that this `danceability` feature, which has so much uncertainty cooked into it, is really capturing something like visceral-ness. We can capture something about Antonoff's music with these variables, and it might just be that subjective "ick" that many people feel and others enjoy and that we're all just trying to describe.

@@

    > I'm literally not an expert, and of course I don't actually claim to always be able to be able to 100% identify a Jack Antonoff song. It's not like I've been practicing the way that I can tell if he's produced something. It's just basically if it sounds to me like he's produced it. I know that's annoying I will clarify first, I think it's mostly in the vocal treatment, that's the thing that is the most off-putting to me. But just every decision feels a little bit wrong to me. If I wanted to be cringe I could point out the Juno 6 Bass, the perfectly stacked vocals, the Juno 6 pad, the wide and tight plate reverb, the Juno 6 sword, the goo goo gaga, the man hates treble unless it's just random crap swishing around, everything gets low passed with thin 707 drum samples, thin 808 drum samples, then 909 drum samples, thin Lin drum samples, thin thin drum samples, all the resonances, the harmonics get removed, everything is sort of squished down to the fundamental reduction in timbre, everything exists in a small frequency range, everything becomes a tone, it's an extremely consonant sound with no interplay between the different elements. Now that is a skill to do, and it produces a cleaner sound, and the only problem is I viscerally hate that sound.

@@

He was really nice and fun to talk to. I bolded the part about vocals sounding the way they would to the person who's singing them while editing this paragraph because I thought it was kind of funny, and maybe meaningful. It sounds like it's coming from inside the house!

@@

I joined these data sets using the `rbind` function in base R, and created an additional variable called "jack", which contains a 1 if Jack Antonoff produced or co-produced the song, and a 0 if he had nothing to do with it. I got the information on which songs Antonoff did or did not produce from the "Jack Antonoff Production Discography" Wikipedia page [@citeWiki].

@@

[^3]: \-- a word I just learned, which refers to "the lowest regular pulse train that a listener intuitively infers from the timing of perceived musical events". The full documentation for each variable included in this study is available in the Appendix

This is pretty vague. Especially valence -- is the positivity related to lyrics, or something else? I compared the valence scores of classical music tracks to see if this score is based on lyrics, or the audio itself. In @tbl-Holst, Gustav Holst's Jupiter has a valence score 326.78% higher than Mars. This tracks. It suggests the valence is based on audio features other than lyrical content. It also suggests that it is, in fact, capturing a difference between the two songs' profiles.

@@

In @fig-instrumentalness, St. Vincent is the only plot with any significant trend.

@@

Finally, in @fig-tempo, there aren't many significant trends, except for Antonoff's collaborations with Lana Del Rey and Taylor Swift.

@@

Bracketing the major blindspot that there variables provide a single number to represent the dynamics of an entire song, the model has a few other limitations. First of all, this model is that assumes that the relationship between the independent variables and the dependent variable is linear. There could be non-linear relationships between the variables that this model isn't capturing. Second, it assumes that the relationship between the independent variables and the dependent variable is the same for all observations, which might not be true in practice -- there could be interactions between the variables, or in certain subgroups of data where the relationship is different. Albums where Jack may have had more of a back seat producing role are one example. The fact that a song's development, from inception to recording, is a whole creative process, each having its own origin story with varying degrees of Jack-ness, is another.

@@

I used **`rank_results()`** from the **`yardstick`** package to rank the models based on accuracy, log loss, sensitivity, and specificity. \[ask rohan about this\]

Downsampled it is.

@@

The model's accuracy score is 0.743 accuracy, and the ROC AUC score is 0.841. The accuracy score could be misleading, since the model had imbalanced classes, and downsampling may have affected the model's ability to correctly identify the minority class. The ROC AUC score is more reliable, since it's less sensitive to class imbalance.

@@

Using the Spotify Audio Features, you can't really gauge what Jack Antonoff's music sound like. You get a rough idea, but you don't get the same granularity that you would by simply DM'ing Caleb Gamman on Twitter and asking him to describe it you. Or by watching the video where he already did that.

@@

The Spotify Features also wouldn't help you understand in the same way you end up understanding it when you sit on my friend Matthew Jordan's couch and watch him beat box a few different Jack Antonoff and Pharrell impressions in order to show the contrast.

@@

[^5]: I encourage you to follow along and drop me an email with your own evaluation at michaela.drouillard\@mail.utoronto.ca.

@@

It's just his subjective taste, but in one way Gamman was actually totally on the money about this. The danceability variable, which is the strongest predictor of Jack-ness, seems to actually be more about "visceral"-ness when you do a close-listen of the songs with the highest danceability scores. The reason Jack is getting under people's skin could just be because his music... gets underneath your skin. Right to the viscera.

@@

One fundamental weakness is cooked into the dataset. Since we're not in the studio, we can't know the degree of Jack-iness in each individual song. Especially when he takes on the role of a more backseat co-producer, as he did with Lana Del Rey's Did You Know There's a Tunnel Under Ocean Blvd? \[tk\]. We can't actually know what his involvement in each song was. There's also the issue that a study like this implies that Jack Antonoff is the reason for the way these artists' sounds develop, which takes that agency away from the artists. This is critique you see on the internet, and also a critique that some people on Twitter leveraged against Gamman after his Midnights video went viral. I mentioned it to a friend when she asked me how my final paper was going, and she said something like: "Yeah but like, it's *Taylor Swift*. As if anyone actually thinks that Jack Antonoff is the reason for her success." I agree.

@@

\appendix

# Additional details

>>>>

---
title: "We Are All Trying to Describe Something And We Don't Have the Words"
subtitle: "A Study of The Jack Antonoffification of Pop Music, and Propagating Uncertainty with Spotify"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/jack."
date: today
abstract: "Using data from the Spotify API, I explored Jack Antonoff's production style and assessed whether Spotify's audio features can capture it. In an interview with Glen McDonald, an engineer at Spotify, I learned that the danceability and valence features were created by having interns tag songs as positive or gloomy, or danceable or un-danceable. Using a logistic regression model, I found that danceability was the most important feature in predicting Antonoff's involvement in a song. In the discussion section, I do a close-listen of Antonoff's most danceable songs, and argue that the danceability variable is actually capturing something more like a visceral quality to Antonoff's music."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Midnights was produced by Swift and her longtime collaborator Jack Antonoff, who is the subject of this paper, and, I suspect, the reason behind both my glitching, and the stranger's "Exactly".

Antonoff is prolific, to the extent that it is starting to bother people. He has produced and co-produced music for Lorde, Taylor Swift, Florence and the Machine, The Chicks, Clairo, The 1975, Grimes, Zayn, Pink, Lana Del Rey, Olivia Rodrigo, the Minions: The Rise of Gru Soundtrack, and more. All of this and his own bands too (Bleachers, and fun.). When I was getting brunch with a friend one Sunday and a Swift song came on, and she said "Jack Antonoff's gonna kill us all", we both laughed. What my friend was referring to is the fact that Jack Antonoff and his sound seems to be everywhere these days. As Andrew Marantz writes in the New Yorker: "When his band releases an album, the world responds politely. When he produces one by Lorde or Lana Del Rey or Taylor Swift, the world wobbles on its axis" [@citeMarantz].

@@

I spoke to Caleb Gamman, a Youtuber and Antonoff-production-style hater who had a tweet go viral around the time of Taylor Swift's *Midnights* release, in which he predicts whether or not a song was produced by Antonoff within seconds. A real-life logistic regression, if you will. Using an actual logistic regression model, I was able to determine that Spotify's `danceability` feature was the most important in predicting whether or not a song had any Antonoff involvement. During an interview with Glen McDonald, an engineer at Spotify, I learned that the `valence` and `danceability` variables were created by hiring college interns and having them tag whether or not songs were positive or danceable. In the discussion section, I listen to the songs with the highest danceability scores, and make an argument that this `danceability` feature, which has so much uncertainty cooked into it, is really capturing something like visceral-ness. Against the odds, we can capture something about Antonoff's music with these variables, and it might just be that subjective "ick" that some people feel, and others enjoy, and that we're all just trying to describe.

@@

    > ...I'm literally not an expert, and of course I don't actually claim to always be able to be able to 100% identify a Jack Antonoff song. It's not like I've been practicing the way that I can tell if he's produced something. It's just basically if it sounds to me like he's produced it. I know that's annoying I will clarify first, I think it's mostly in the vocal treatment, that's the thing that is the most off-putting to me. But just every decision feels a little bit wrong to me. If I wanted to be cringe I could point out the Juno 6 Bass, the perfectly stacked vocals, the Juno 6 pad, the wide and tight plate reverb, the Juno 6 sword, the goo goo gaga, the man hates treble unless it's just random crap swishing around, everything gets low passed with thin 707 drum samples, thin 808 drum samples, then 909 drum samples, thin Lin drum samples, thin thin drum samples, all the resonances, the harmonics get removed, everything is sort of squished down to the fundamental reduction in timbre, everything exists in a small frequency range, everything becomes a tone, it's an extremely consonant sound with no interplay between the different elements. Now that is a skill to do, and it produces a cleaner sound, and the only problem is I viscerally hate that sound. ...

@@

He was really nice and fun to talk to. I bolded the part about vocals sounding the way they would to the person who's singing them while editing this paragraph because I thought it was kind of funny, and kind of cool. It's coming from inside the house!!!!

@@

I joined these data sets using the `rbind` function in base R, and created an additional variable called "jack", which contains a 1 if Jack Antonoff produced or co-produced the song, and a 0 if he had nothing to do with it. I got the information on which songs Antonoff did or did not produce from the "Jack Antonoff Production Discography" Wikipedia page [@citewiki].

@@

[^3]: \-- a word I just learned, which refers to "the lowest regular pulse train that a listener intuitively infers from the timing of perceived musical events".

This is pretty vague. Especially valence -- is the positivity related to lyrics, or something else? I compared the valence scores of classical music tracks to see if this score is based on lyrics, or the just the instruments. In @tbl-Holst, Gustav Holst's Jupiter has a valence score 326.78% higher than Mars. This tracks. It suggests the valence is based on audio features other than lyrical content. It also suggests that it is, in fact, capturing a difference between the two songs' profiles.

@@

In @fig-instrumentalness, St. Vincent is the only plot with any significant trend in instrumentalness.

@@

Finally, in @fig-tempo, there aren't many significant trends in tempo, except for Antonoff's collaborations with Lana Del Rey and Taylor Swift.

@@

Bracketing the major blindspot that there variables provide a single number to represent the dynamics of an entire song, the model has a few other limitations. First of all, this model assumes that the relationship between the independent variables and the dependent variable is linear. There could be non-linear relationships between the variables that this model isn't capturing. Second, it assumes that the relationship between the independent variables and the dependent variable is the same for all observations, which might not be true in practice -- there could be interactions between the variables, or in certain subgroups of data where the relationship is different. Albums where Jack may have had more of a back seat producing role are one example. (The fact that a song's development, from inception to recording, is a whole creative process, each having its own origin story with varying degrees of Jack-ness, is another.)

@@

I used **`rank_results()`** from the **`yardstick`** package to rank the models based on accuracy, log loss, sensitivity, and specificity, and went with the downsampled model.

@@

The model's accuracy score is 0.743, and the ROC AUC score is 0.841. The accuracy score could be misleading, since the model had imbalanced classes, and downsampling may have affected the model's ability to correctly identify the minority class. The ROC AUC score is more reliable, since it's less sensitive to class imbalance.

@@

Using the Spotify Audio Features, you can't really gauge what Jack Antonoff's music *sounds* like. You get a rough idea, but you don't get the same granularity that you would by simply DM'ing Caleb Gamman on Twitter and asking him to describe it you. Or by watching the video where he already did that.

@@

The Spotify Features also wouldn't help you understand in the same way you end up understanding it if you sat on my friend Matthew Jordan's couch and watched him beat box a few different Jack Antonoff and Pharrell impressions in order to show the contrast.

@@

[^5]: I encourage you to follow along and drop me an email with your own assessment at michaela.drouillard\@mail.utoronto.ca.

@@

It's just his subjective taste, but in one way Gamman was actually totally on the money about this. The danceability variable, which is the strongest predictor of Jack-ness, seems to actually be more about "visceral"-ness when you do a close-listen of the songs with the highest danceability scores. Not all of them are danceable, but the one's that aren't still feel like they're trying to manipulate the way your body is holding itself in some way. The reason Jack is getting under people's skin could just be because his music... gets underneath your skin. Right to the viscera.

@@

One fundamental weakness is cooked into the dataset. Since we're not in the studio, we can't know the degree of Jack-iness in each individual song. Especially when he takes on the role of a more backseat co-producer. We can't actually know what his involvement in each song was. There's also the issue that a study like this implies that Jack Antonoff is the reason for the way these artists' sounds develop, which takes that agency away from the artists. This is a critique you see on the internet, and also a critique that some people on Twitter leveraged against Gamman after his Midnights video went viral. I mentioned it to a friend when she asked me how my final paper was going, and she said something like: "Yeah but like, it's *Taylor Swift*. As if anyone actually thinks that Jack Antonoff is the reason for her success." I agree.

@@
